A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex

\newcommand\xend{{\hfill\footnotesize$\blacksquare$}}

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable ${\Varec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = {\Varec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following rule exhaustively:
\[
\hbox{\(
\inferrule{
  {\Varec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ {\Varec{x}} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}
}
\)
\rn{Update}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{y}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}}, {\Varec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ {\Varec{x}} \mapsto {\Varec{\const{C}(y,x)}}, {\Varec{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
are equivalent for any value of ${\Varec{y}}$.
\xend
\end{example}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Codata
  \quad
  \left|\tau\right| = 1
  \quad
  t, u : \tau \in \tEc
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc
  \quad
  ( s( t ) \in \tEc \text{ and } s \in \Sel^\tau )
  \text{ or }
  ( \tau \in \Data \text{ and } \tau \text{ finite} )
}{
  \Ec := \Ec,\: t \teq \const{C}_1(\const s^1_1( t ), \ldots,\const s^{n_1}_1( t ) ) \ror \cdots \ror \Ec := \Ec,\: t \teq \const{C}_m(\const s^1_m( t ), \ldots,\const s^{n_m}_m( t ) )
}
\)
\rn{Split}
\end{tabular}
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

>>>>

%%% @ANDY: I don't like using \qed or anything too similar (e.g. \Box) for
%%% anything else than "quid erat demonstrandum". I'm open to other symbols,
%%% e.g. a filled black square. But if the example are short enough, which
%%% they currently are, we could live with nothing.
%%%
%\newcommand\xend{{\hfill\footnotesize$\blacksquare$}}
%\newcommand\xend{{\qed}}
\newcommand\xend{{}}

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable ${\Varec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = {\Varec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following unfolding rule exhaustively:
\[
\hbox{\(
\inferrule{
  {\Varec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ {\Varec{x}} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}
}
\)
\rn{Unfold}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Unfold} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Unfold $\Varec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Unfold $\Varec{y}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{example}

@@

%% @ANDY: Minor terminology point:
%% A function symbol can have a "type signature" (but we're trying to avoid that terminology);
%% unless it's a constant, it doesn't have a (first-order) type. So I'm rephrasing.

\begin{example}
Let $\const{C} : $ $\tau_1 \times \tau_2 \rightarrow \tau_2$ be a constructor,
and let $\Ec = \{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
Constructing $\Val$ as described above yields
\[\begin{array}{rcrlrl}
\Val \ec{x} & \,=\, &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, & \mu {\VAREC{\const{C}(y, x)}}. & \const{C}( {\Varec{y}},\, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\VAREC{\const{C}(y,x)}}. & \const{C}( {\Varec{y}},\, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, {\VAREC{\const{C}(y,x)}} ) )
\end{array}\]
Since $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are $\alpha$-equivalent,
%for the renaming
%$\{ {\Varec{x}} \mapsto {\VAREC{\const{C}(y,x)}}, {\VAREC{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
the calculus can infer $x \teq \const{C}(y,x)$.
%Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
%are equivalent for any value of ${\Varec{y}}$.
\xend
\end{example}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Codata
  \quad
  \left|\tau\right| = 1
  \quad
  t, u : \tau \in \tEc
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc
  \quad
  ( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\tau )
  \text{ or }
  ( \tau \in \Data \text{ and } \tau \text{ finite} )
}{
  \Ec := \Ec,\: t \teq \const{C}_1(\const s^1_1( t ), \ldots,\const s^{n_1}_1( t ) ) \ror \cdots \ror \Ec := \Ec,\: t \teq \const{C}_m(\const s^1_m( t ), \ldots,\const s^{n_m}_m( t ) )
}
\)
\rn{Split}
\end{tabular}
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}
